---
title: Shiny
author: Jonathan Marshall
date: 12 May 2016
output: 
  ioslides_presentation: 
    highlight: tango
---

```{r, setup, echo=FALSE}
library(knitr)
library(shiny)
knitr::opts_chunk$set(echo=FALSE, fig.width=8, dev.args=list(bg='transparent'), comment="")
```

# What is Shiny?

## What is Shiny?

- Shiny is an R package.

- It includes a web-server that allows linking to R.

- This allows interactive web applications (built with HTML, CSS + Javascript) to
run R code.

- Also allows the R code to describe and update the application/interface.

## The R side

- Shiny applications are usually built using only R-code.

- There are a bunch of R functions that build up the HTML that
defines the interface (widgets, plots, text etc.)

- In addition, R functions are defined that respond to changes
in the interface (e.g. clicks, or values selected in widgets).

## Defining the interface: `ui.R`

- The interface is normally defined in `ui.R`.

- The function to do so is `shinyUI`, which takes as it's argument
a composition of functions that define the layout.

- These functions are basically R wrappers that produce HTML
code (`<div>` etc.)

- Most of these use the `bootstrap` javascript library to define
layouts and widgets.

## Example

```{r, echo=TRUE, eval=FALSE}
shinyUI(fluidPage(
  titlePanel("Hello Shiny!"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
                  "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)
    ),
    mainPanel(
      plotOutput("distPlot")
    )
  )
))
```

## Example

- There is a `fluidPage` which creates a page layout that includes rows which
in turn include columns.

- Fluid in the sense that the width/height of the browser is taken into account
during layout.

- You then typically include `fluidRow`'s within the page that define elements
that appear on the same row.

## Example

- You can also include higher-level layout tools, such as `sidebarLayout` which splits the
page into left and right panes.

- The left pane has a distinct background colour and is usually where you place input widgets/controls.
You usually use a `sidebarPanel` for this.

- The right pane is larger (2/3rds the width) and is where you'd normally place output. You usually
use a `mainPanel` for this.

## Example

- Within the `sidebarPanel` and `mainPanel` you can then include other stuff.

- e.g. the `sliderInput` for a slider (this is an input widget).

- e.g. the `plotOutput` control where plots will end up (this is an output widget).

## Defining the app: `server.R`

- What the app does is defined in `server.R`.

- This contains the function `shinyServer` which is passed
a function that takes `input` and `output` as parameters.

- `input` contains information from the interface widgets (as a list).

- `output` contains information to pass to output widgets (e.g. plots).

## Example

```{r, echo=TRUE, eval=FALSE}
shinyServer(function(input, output) {
  output$distPlot <- renderPlot({
    x    <- faithful[, 2]  # Old Faithful Geyser data
    bins <- seq(min(x), max(x), length.out = input$bins + 1)

    # draw the histogram with the specified number of bins
    hist(x, breaks = bins, col = 'darkgray', border = 'white')
  })
})
```

## Example

- Notice the `shinyServer` function is passed afunction
of `input` and `output`.

- The inner-most code contains some expressions for drawing a histogram.

- This expression is wrapped in the `renderPlot` function.

- The output from `renderPlot` is placed in the `distPlot` member of
`output`.

## Example: How it all works.

- In `ui.R` we defined two controls:
    - The sliderInput, which took first parameter `bins`.
    - The plotOutput, which took first parameter `distPlot`.

- In `server.R` then `plotRender` function took an expression that used
`input$bins` and set `output$distPlot`.

- The `plotRender` function is **reactive**. It re-executes automatically
whenever any of the things it uses (e.g. `input$bins`) change.

- Thus, `distPlot` is automatically updated whenever `input$bins` changes,
which is what gives the interactive nature.

## Structure of an R shiny app

- Just a directory.

- Containing `ui.R` and `server.R`.

- RStudio automatically recognises this, and adds a **Run App* icon to
the toolbar.

- Clicking the button is equivalent to running `shiny::runApp` on that folder.

## Have a play!

- Copy the folder `example1`.

- Load up either `ui.R` or `server.R`.

- Hit `Run App`.

- Try:
    - Changing the title.
    - Changing the minimum value for the slider bar.
    - Change the histogram colour.

## User interface elements

- The easiest to start with are the ones we've seen: `titlePanel` and `sidebarLayout`.

- Inside the `*Panel` functions, you can place basically any content that you like: essentially
all the functions produce HTML.

- Most of the common HTML5 tags have equivalents:
    - `p()` is the same as `<p>` for a paragraph of text.
    - `a()` is the same as `<a>` for a hyper-link.
    - `div()` is the same as `<div>` for a division of text etc.
    - `HTML` can be used to pass a character string of HTML code.

## Headers

- Use `h1` through `h6` for first through sixth-level headers (equivalent to single through six
hashes in markdown).

```{r, echo=TRUE, eval=FALSE}
shinyUI(fluidPage(
  titlePanel("A title bar"),
  sidebarLayout(
    sidebarPanel(),
    mainPanel(
      h1("First level title"),
      h2("Second level title"),
      h3("Third level title"),
      h4("Fourth level title")
    )
  )
))
```

## HTML functions

- All the `h1` etc. functions (and most of the HTML generating functions) take additional
arguments.

- Named arguments become attributes in the HTML (e.g. `align='center'` to align `h1` tags)

- Un-named arguments become children in the HTML. These can be tags themselves (i.e. other
HTML functions such as `p()`) or plain-text which become text nodes in the HTML.

```{r}
h1("test", align="center")
p("A paragraph containing", strong("bold"), "blue text.")
```

## HTML functions

- `div` is used for a division of text (basically elements that have a given style).

- `span` is the equivalent of `div` but is contained in-line an existing element.

- `br` can be used for a line break (blank line).

- `em` and `strong` for italics and bold.

- `code` for a code block (mono-spaced font on different background).

- `img` for adding images.

- `tag` for creating arbitrary HTML5 tags that don't have an existing R function.

## Have a play!

- Try creating an app with a title, sidebar and main panel.

- Have a bit of writing in the main panel (a header, some styled text).

- Pop some example R code in as well formatted up as code.

- Maybe add an image, or try changing the colour or alignment of text.

## Widgets

- All input widgets have an identifier (the name of the widget) and a label
(that appears in your app).

- In addition, they take parameters that define their scope, though most
of these have defaults.

- You can have sliders, buttons, select boxes (combobox) text or numeric
input fields, date and range selectors, help text, radio buttons, checkboxes
etc.

- The functions are usually named `*Input()` or `*Button`.

## Try out the widgets

- Grab the `widgets` folder.

- Load up the `ui.R`.

- Run the app and play around.

- Take a look through `ui.R` and see how the widgets are constructed.

## Reactive output

- Output that reacts to your widgets are defined by:

    - placing an R `output` object in your `ui.R`.
    - adding code to set the content of that object to `server.R`.

- The output objects are `htmlOutput`, `imageOutput`, `plotOutput`,
`tableOutput`, `textOutput`, and `uiOutput`.

- What they contain are reasonably self-explanatory!

## Reactive output

- Each of these have a corresponding reactive function that renders
the content of those output controls.

- `renderPlot` produces a plot, `renderImage` produces an image (saved link to a file),
`renderTable` produces a table etc.

- These all take an R expression (surrounded by braces `{}`). Shiny runs the expression
the first time the app is run, then re-runs them as-needed whenever the inputs change.

## Example

- Load up the `reactive/ui.R`. Add in the following `server.R`

```{r, eval=FALSE, echo=TRUE}
shinyServer(function(input, output) {
     output$text1 <- renderText({ 
          "You have selected this"
     })
  }
)
```

- When you run it you'll see the text but it doesn't ever change, no matter
what you do.

- To make it reactive, we need to give it something that changes, i.e. something in `input`.

## Example

- Modify `server.R` to be

```{r, eval=FALSE, echo=TRUE}
shinyServer(function(input, output) {
     output$text1 <- renderText({ 
          paste("You have selected", input$var)
     })
  }
)
```

- Run the app now. You should see that when you change the select input, the output changes as well.

- Add another line of text to the UI and hook it up to react to the slider input.

## Structuring `server.R`

- There are basically three places to put R code in `server.R`.

- Code placed before `shinyServer` will be run once only when your app loads. This is useful for setup code
such as reading in data, loading packages etc.

- Code placed inside `shinyServer` but outside a `render` function (or other reactive function)
is run once whenever a user goes to your app (e.g. refreshing it in a browser).

- Code placed inside `render` (or other reactive functions) runs whenever an input function it is using
changes.

## Example: using data and plots

- Load up the `ui.R` and `server.R` scripts in the `census` folder.

- You'll need to install the `maps` and `mapproj` packages.

- Try running the app. You should get a blue map of the US.

- Look through `server.R` to see how it is using data and some
other R code.

## Try making this reactive

- Try hooking up the `var` select input to change the legend title.

- Try also changing the `min` and `max` to use the slider values. Note that
the range slider produces a vector of length 2 (one for the left and one
for the right).

- Try hooking up the `var` select input to change the data used for the plot. The `switch` function
might be useful here.

- Try changing colour based on the select input, or alternatively, try adding a colour select input.

## Some debugging help!

- A simple way to debug what is going on in shiny apps is to trace what is
happening with different `input` values.

- You can do this using `cat` (or `print`) as long as you are inside a reactive function
when you do it.

- If you are outside a reactive function, you'll find these don't work.

- e.g. try adding `print(input$var)` inside the `renderPlot` function in the census app.

- Now try adding it before the `renderPlot` function (but inside the `shinyServer` function).

- Lastly, try wrapping it with `isolate` using `print(isolate(input$var))`.

## Reactive functions

- By default, the `render` functions are reactive, in that any change in the inputs that they
use will mean re-running the expression.

- This means your expressions need to be quite lean: They need to change only as much as you need
in order to update the interface.

- e.g. you don't want changing the colour of a plot to re-invoke all the processing code that was needed
to create the data for the plot.

- So it pays to break your code up so that the slow bits only run when they have to.

## Example: stockVis

- This uses financial data from yahoo.

- You'll need to install the `quantmod` package first.

- Load up the `stockvis` app and run it.

- Type in a stock symbol (e.g. GOOG, AAPL, GS)

- Try clicking on the `Plot y axis on log scale` button.

## stockVis app

- This app has a problem: Each time you toggle the log scale button, the `renderPlot` command runs.

- This in-turn means `getSymbols` is run, which re-fetches the data from Yahoo. This is slow!

- We can remedy this by breaking the function up into one that fetches the data and one that plots.

- We use the `reactive` function for this.

## Example: stockVis v2

Try changing the `shinyServer` function to:
```{r, eval=FALSE, echo=TRUE}
dataInput <- reactive({
  getSymbols(input$symb, src = "yahoo", 
    from = input$dates[1],
    to = input$dates[2],
    auto.assign = FALSE)
})
output$plot <- renderPlot({    
  chartSeries(dataInput(), theme = chartTheme("white"), 
    type = "line", log.scale = input$log, TA = NULL)
})
```

## Reactive functions

- They're smarter! Whenever they're run, they cache their output value.

- Next time they're run, they check whether the cached output value is out of date. If not, it just
returns the cached value.

- It knows whether it's out of date by monitoring it's input values.

- So `dataInput` will only be computed if `input$symb` or `input$dates` change.

- The neat thing is that `renderPlot` uses `dataInput`, so it will in-turn change only if `dataInput`
or `input$log` changes.

- Changing only `input$log` thus just re-plots, it doesn't download the data.

## Showcase mode

- You can see which bits of code are running while you play with the app by switching to showcase mode.

- Run the app using

```{r, echo=TRUE, eval=FALSE}
runApp("stockvis", display.mode = 'showcase')
```

- You'll notice which bits of the `server.R` code run when you interact with the interface.

## Hooking up inflation adjustment.

- Change the `renderPlot` statement to:

```{r, eval=FALSE, echo=TRUE}
output$plot <- renderPlot({   
    data <- dataInput()
    if (input$adjust) data <- adjust(dataInput())
    
    chartSeries(data, theme = chartTheme("white"), 
        type = "line", log.scale = input$log, TA = NULL)
  })
```

- This will again be problematic when you toggle `input$log`.

- Try fixing it!

## Deploying apps

There are three options for deploying apps:

1. Give users the app and Rstudio. They run it themselves from RStudio. It's free.

2. Deploy the app using a server owned by someone else (e.g. shinyapps.io by rstudio). User just needs a browser. You get a small amount of bandwidth/users for free, else you can pay for larger hosting options.

3. Run a Shiny server on your network. This is usually a *nix server that runs R as well as a webserver. It's free if you maintain it yourselves. RStudio provides a support for $.

## Adding apps to documents

- As the app is just HTML/CSS/javascript, then you can incorporate into a markdown document.

- The document must then be run from shiny though for the interactive bits to work.

- Useful for presentations that you run from RStudio.

- Can also be used if you link to the app (e.g. via `<iframe>`) on a shinyserver.
