---
title: Spatial statistics
author: Jonathan Marshall
date: 17 June 2016
output: 
  ioslides_presentation: 
    highlight: tango
---

# Overview

## Overview

- Data types (Points vs Areal)
- Mapping
- Coordinate transformation
- Point process data
- Spatial models

## Data types

There are two main types of spatial (or spatio-temporal) data.

- **Points**, where locations of data are given in long/lat or some other coordinate system.
- **Areal**, where counts are aggregated over regions (countries, counties etc.) or values are
given for a region (e.g. GDP by country).

The type of data affects how it is modelled (and how it's mapped, but the latter
is more obvious!)

## Point data

- Each observation has a location (and time).
- May also have additional value information (e.g. elevation, case vs control indicator).
- We normally want to do things like map where the points are.
- And maybe model whether the points are spatially clustered or correlate with some
other spatial or temporal data.
- Usually a data frame is enough to store these data.

## Areal data

- Each observation defines a region (usually a polygon boundary describing shape and location).
- In addition have a number of covariates for each region.
- Need to store a polygon boundary (or list of polygons) for each region plus
a data frame for the covariates.
- We usually use a `SpatialPolygonsDataFrame` from the `sp` package for this.

## Mapping

- Data for a map consist of points, lines, polygons, and rasters.
- A line is just a collection of joined points.
- A polygon is just a collection of joined points where the start and end points are the same.
- A raster is just a collection of values defined over a uniform grid of points.
- Key thing when mapping is the background is important. It gives spatial context to the data.

## Drawing maps

Lots of packages are available for drawing maps in R from a variety of sources.

- `sp` is the spatial package used for drawing things.
- `maptools` can handle loading ESRI shapefiles (common format).
- `rgeos` can handle loading shapefiles that have projections in them.
- `dismo` and `RgoogleMaps` can be used for google maps.
- `Rworldmap` contains maps of the world.
- `leaflet` can be used for visualising spatial data in shiny apps (using openstreetmap maps)
- `googleVis` can be used for interactive maps using the Google visualisation API

## Shapefiles

- ESRI shapefiles are one of the most readily available geospatial data.
- Supported in a bunch of GIS packages.
- Most countries supply data in this format.
- Can get from [stats NZ](http://www.stats.govt.nz/browse_for_stats/Maps_and_geography/Geographic-areas/digital-boundary-files.aspx)

## Areal units in NZ

- There are a variety of boundaries for various adminstrative entities in NZ.
- The lowest level of spatial areal data in NZ is the meshblock (MB).
- These are arranged in area units (AU).
- Which are arranged in territorial authorities (TA).
- There's also regional councils, wards, electorates, public health areas etc.
- Many of these don't have a 1:1 mapping. e.g. MidCentral DHB area vs MidCentral Public Health Unit.
- Typically meshblocks have a many to 1 map with each of the other boundaries.

## Example: Midcentral DHB

```{r, echo=TRUE, eval=FALSE}
library(sp)
library(maptools)
midcentral <- readShapeSpatial("shape/midcentral.shp")
plot(midcentral)
```

## Example: Midcentral DHB

```{r, echo=FALSE}
library(sp)
library(maptools)
midcentral <- readShapeSpatial("shape/midcentral.shp")
plot(midcentral)
```

## Example: Midcentral DHB

```{r}
names(midcentral)
hist(midcentral$POPULATION)
```

## Example: Midcentral DHB

Seems there's something wrong with the `AREA` column
```{r}
summary(midcentral$AREA)
```
But we have this information in the polygons
```{r}
midcentral@polygons[[1]]@area
```

## Extracting information from polygons

The `SpatialPolygonsDataFrame` has a number of `slots`, one of which is `polygons`
and the other is `data`. There's also some other things.
```{r}
slotNames(midcentral)
```
We can extract slots using the `@` operator, or using `slot`
```{r}
class(midcentral@polygons)
class(slot(midcentral, 'polygons'))
```

## Extracting information from polygons

To extract information from the polygons, we can use `lapply` (list apply)
on the list to pull out what we want using the `slot` command:

```{r}
areas <- lapply(midcentral@polygons, slot, 'area')
head(areas, n=2)
```

## Extracting information from polygons

Can convert this to a vector using `unlist`

```{r}
areas <- unlist(lapply(midcentral@polygons, slot, 'area'))
midcentral$AREA <- areas
summary(midcentral$AREA)
```

## Plotting information on the map

Say we want to colour the meshblocks by their area. To do this
we need to assign colours to the various areas. We do this by binning
the data and then assigning each bin a colour.

```{r, eval=FALSE}
cols <- cut(midcentral$AREA, quantile(midcentral$AREA))
plot(midcentral, col=cols)
```

## Plotting information on the map

```{r, echo=FALSE}
cols <- cut(midcentral$AREA, quantile(midcentral$AREA))
plot(midcentral, col=cols)
```

## Plotting information on the map

```{r, eval=FALSE}
bins <- cut(midcentral$AREA, quantile(midcentral$AREA))
cols <- heat.colors(5)[bins]
plot(midcentral, col=cols, border=NA)
```

## Plotting information on the map

```{r, echo=FALSE}
bins <- cut(midcentral$AREA, quantile(midcentral$AREA))
cols <- heat.colors(5)[bins]
plot(midcentral, col=cols, border=NA)
```

## Plotting information on the map

```{r, eval=FALSE}
library(RColorBrewer)
bins <- cut(midcentral$AREA, quantile(midcentral$AREA))
cols <- rev(brewer.pal(5, "BuPu"))[bins]
plot(midcentral, col=cols, border="lightgrey", lwd=0.1)
```

## Plotting information on the map

```{r, echo=FALSE}
library(RColorBrewer)
bins <- cut(midcentral$AREA, quantile(midcentral$AREA))
cols <- rev(brewer.pal(5, "BuPu"))[bins]
plot(midcentral, col=cols, border="lightgrey", lwd=0.1)
```

## Colour choice

- Choosing the bins and the colour can make a large difference
to the overall narrative taken from the map.
- In general around 5-7 shades is normally good.
- Colour scales should be appropriate
    - a gradient for quantitative variables
    - divergent gradients for things where positive/negative differ
    - non-sequential colours for categories.
- [Bad colour choice](http://weather.graphics/nbm/blend.php)

## Using `classInt` for discretising numeric variables

```{r, eval=FALSE, echo=TRUE}
library(classInt)
bins <- classIntervals(midcentral$AREA, n=7, style="jenks")
cols <- findColours(bins, c("red", "blue"))
plot(midcentral, col=cols, lwd=0.2, border="lightgrey")
```

## Using `classInt` for discretising numeric variables

```{r, echo=FALSE}
library(classInt)
bins <- classIntervals(midcentral$AREA, n=7, style="jenks")
cols <- findColours(bins, c("red", "blue"))
plot(midcentral, col=cols, lwd=0.2, border="lightgrey")
```

## Try it yourself

- Create a map of midcentral
- Colour using the `POPULATION` variable.
- Try different `style` arguments to `classIntervals`.

## Using google maps

```{r, echo=TRUE, eval=FALSE}
library(RgoogleMaps)

map = GetMap(center = c(lat=-40.3397633, lon=175.6212577), zoom=16)
PlotOnStaticMap(map)
```

## Using google maps

```{r, echo=FALSE}
library(RgoogleMaps)

map = GetMap(center = c(lat=-40.3397633, lon=175.6212577), zoom=16)
PlotOnStaticMap(map)
```

## Using google maps: Adding points

```{r, eval=FALSE}
library(RgoogleMaps)

map = GetMap(center = c(lat=-40.3397633, lon=175.6212577), zoom=16)
zombies = data.frame(lat=rnorm(100, -40.339, 0.002), long=rnorm(100, 175.622, 0.003), case=rbinom(100, 1, 0.2)+1)
PlotOnStaticMap(map, zombies$lat, zombies$lon, col=zombies$case)
```

## Using google maps: Adding points

```{r, echo=FALSE}
library(RgoogleMaps)

map = GetMap(center = c(lat=-40.3397633, lon=175.6212577), zoom=16, maptype="satellite")
zombies = data.frame(lat=rnorm(100, -40.339, 0.002), long=rnorm(100, 175.622, 0.003), case=rbinom(100, 1, 0.2)+1)
PlotOnStaticMap(map, zombies$lat, zombies$lon, col=zombies$case, pch=19)
```

## Using googleVis

```{r, echo=TRUE, eval=FALSE}
library(googleVis)
zombies$latlon <- paste(zombies$lat, zombies$lon, sep=":")
zombies$text   <- ifelse(zombies$case == 2, "zombie", "normal")
map.gv <- gvisMap(zombies, locationvar="latlon", tipvar="text",
                  options = list(showTip=T, showLine=F, enableScrollWheel=TRUE,
                           useMapTypeControl=T, width=1400,height=800))
plot(map.gv)
```

## Using googleVis

```{r, echo=FALSE}
library(googleVis)
zombies$latlon <- paste(zombies$lat, zombies$lon, sep=":")
zombies$text   <- ifelse(zombies$case == 2, "zombie", "normal")
map.gv <- gvisMap(zombies, locationvar="latlon", tipvar="text",
                  options = list(showTip=T, showLine=F, enableScrollWheel=TRUE,
                           useMapTypeControl=T, width=1400,height=800))
plot(map.gv)
```

## Converting coordinate systems

- All maps come with a projection that maps the two x/y coordinates onto the (not-spherical) earth in some way.
- e.g. the `midcentral` dataset is using new zealand map grid projection.
- e.g. Google Maps uses WGS84 latitude and longitude.
- It's often essential to be able to convert between these.
- `spTransform` can use this, using `proj4` projection schemes.
- They're somewhat archaic!

## Converting coordinate systems

```{r}
library(rdgal)
midcentral <- readOGR("shape/midcentral.shp", layer="midcentral")
midcentral@proj4string
midcentral@polygons[[1]]@labpt
mc.lonlat <- spTransform(midcentral, CRS = "+proj=longlat +datum=WGS84")
mc.lonlat@proj4string
mc.lonlat@polygons[[1]]@labpt
```

## Converting coordinate systems

```{r, eval=FALSE}
mb_centers <- simplify2array(lapply(mc.lonlat@polygons, slot, 'labpt'))
mb_centers <- data.frame(lon = mb_centers[1,], lat=mb_centers[2,])
map = GetMap(center = c(lat=-40.3397633, lon=175.6212577), zoom=9, maptype="satellite")
PlotOnStaticMap(map, mb_centers$lat, mb_centers$lon, col="red")
```

## Converting coordinate systems

```{r, echo=FALSE}
mb_centers <- simplify2array(lapply(mc.lonlat@polygons, slot, 'labpt'))
mb_centers <- data.frame(lon = mb_centers[1,], lat=mb_centers[2,])
map = GetMap(center = c(lat=-40.3397633, lon=175.6212577), zoom=9, maptype="satellite")
PlotOnStaticMap(map, mb_centers$lat, mb_centers$lon, col="red")
```

## Plotting polygons on maps

```{r, echo=FALSE}
library(PBSmapping) # needed for colours
map <- GetMap(center = c(lat=-40.3397633, lon=175.6212577), zoom=9, maptype="satellite")
# extract polygons
mc.polys <- as.SpatialPolygons.PolygonsList(mc.lonlat@polygons)
bins <- classIntervals(midcentral$POPULATION, n=7, style="jenks")
cols <- findColours(bins, c("red", "blue"))
PlotPolysOnStaticMap(map, mc.polys, col=cols)
```

## Plotting data from other data frames

- Given a `SpatialPolygonsDataFrame` object that has the geographic regions,
how do we add data from another source.
- e.g. suppose we have another data source that defines the population of the meshblock
in 2013 rather than 2006.
- How do we merge them?
- Can use `dplyr`. Key is that the identifiers need to be the same.

## Plotting data from other data frames

```{r}
library(meshblocknz)
head(mb2006)
```

## Plotting data from other data frames

```{r, echo=TRUE, eval=FALSE}
library(dplyr)
midcentral@data = midcentral@data %>%
  left_join(mb2006, by=c('MB06'='MB2006')) %>%
  mutate(PopGrowth = Pop2013 - Pop2006)
summary(midcentral$PopGrowth)
bins <- classIntervals(midcentral$PopGrowth, n=7, style="fixed", fixedBreaks=c(-300,-30,-10,-2,2,10,30,300))
cols <- findColours(bins, rev(brewer.pal(7, "RdYlBu")))
plot(midcentral, col=cols, border="lightgrey", lwd=0.2)
```

## Plotting data from other data frames

```{r, echo=FALSE}
library(dplyr)
midcentral@data = midcentral@data %>%
  left_join(mb2006) %>%
  mutate(PopGrowth = Pop2013 - Pop2006)
bins <- classIntervals(midcentral$PopGrowth, n=7, style="fixed", fixedBreaks=c(-300,-30,-10,-2,2,10,30,300))
cols <- findColours(bins, rev(brewer.pal(7, "RdYlBu")))
plot(midcentral, col=cols, border="lightgrey", lwd=0.2)
```

## Plotting data from other data frames

```{r, eval=FALSE}
plot(midcentral, col=cols, border="lightgrey", lwd=0.2,
     xlim=c(2726000, 2738000), ylim=c(6088000, 6094000))
```
## Plotting data from other data frames

```{r, echo=FALSE}
plot(midcentral, col=cols, border="lightgrey", lwd=0.2,
     xlim=c(2726000, 2738000), ylim=c(6088000, 6094000))
```